Rank Verification for Exponential Families

### 2016-04-22 verifying-winner.tex
Fixed natbib quirks
Sections shuffled, paper unconcluded

\captionsetup{width=0.8\textwidth}
\DeclareMathOperator*{\argmax}{arg\,max}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\setlength{\tabcolsep}{1em}
\setlist{listparindent=\parindent, parsep=0pt}

\newcommand{\EE}{\mathbb{E}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}

@@

\title{Rank verification for exponential families}
\author{Kenneth Hung \and William Fithian}
\date{\today}
\maketitle

@@

\section{Introduction}
\label{sec:introduction}

\subsection{Motivating Example: Iowa Republican Caucus Poll}\label{sec:iowa}

Table~\ref{tbl:poll} shows the result of a Quinnipiac University poll asking 890 Iowa Republicans their preferred candidate for the Republican presidential nomination \citep{quinnipiac}. Donald Trump led with $31\%$ of the vote, Ted Cruz came second with $24\%$, Marco Rubio third with $17\%$, and ten other candidates including ``Don't know'' trailed behind.

@@

Seeing that Trump leads this poll, several salient questions may occur to us: Is Trump really winning, and if so by how much? And analogously, is Cruz really in second, is Rubio really in third, and so on? Note that there is implicitly a problem of multiple comparisons here, because if Cruz had led the poll instead, we would be asking a different set of questions. Indeed, the selection issue appears especially pernicious due to the so-called ``winner's curse:'' given that Trump leads the poll, it more likely than not overestimates his support.

Nevertheless, if we blithely ignore the selection issue, we might carry out the following analyses to answer the questions we posed before at significance level $\alpha=0.05$. We assume for simplicity that the poll represents a simple random sample of Iowa Republicans, i.e., that the data are a multinomial sample of size $890$ and underlying probabilities $(\pi_{\text{Trump}}, \pi_{\text{Cruz}}, \ldots)$.\footnote{The reality is a bit more complicated: before releasing the data, Quinnipiac has processed it to make the reported result more representative of likely caucus-goers.}
\begin{enumerate}
\item {\em Is Trump really winning?} If Trump and Cruz were in fact tied, then Trump's share of their combined 490 votes would be distributed as $\text{Binom}(490, 0.5)$. Because the (two-tailed) $p$-value for this pairwise test is $p=0.006$, we reject the null and conclude that Trump is really winning.
\item {\em By how much?} Using an exact $95\%$ interval for the same binomial model, we conclude Trump has at least $7.5\%$ more support than Cruz (i.e., $\pi_{\text{Trump}} \geq 1.075 \,\pi_{\text{Cruz}}$) and also leads the other candidates by at least as much.
\item {\em Is Cruz in second, Rubio in third, etc.?} We can next compare Cruz to Rubio just as we compared Trump to Cruz (again rejecting because 214 is significantly more than half of 365), then Rubio to Carson, and so on, continuing until we fail to reject. The first four comparisons are all significant at level $0.05$, but Paul and Bush are tied so we stop.
\end{enumerate}

@@

\subsection{Generic Problem Setting and Main Result}

Generically, we will consider data drawn from an exponential family model with density
\begin{equation}\label{eq:expfam}
X \sim e^{\theta'x - \psi(\theta)}g(x),
\end{equation}
with respect to either the Lebesgue or counting measure on $\RR^n$. We assume further that $g(x)$ is symmetric. In addition to the multinomial family, \eqref{eq:expfam} also encompasses settings such as comparing independent binomial treatment outcomes in a clinical trial, competing sports teams under a Bradley-Terry model, entries of a Dirichlet distribution, and many more; see Section~\ref{sec:examples} for these and other examples. \WFcomment{Also students taking a test in a Rasch model, after conditioning on how many get each question right. Or comparing the variances of different normal populations.}

We will generically use the term {\em population} to refer to the treatment group, sports team, political candidate, etc. represented by a given random variable $X_j$. As we will see, $\theta_j \geq \theta_k$ if and only if $X_j$ is stochastically larger than $X_k$; thus, there is a well-defined stochastic ordering of the populations that matches the ordering of the entries of $\theta$. We will refer to the population with maximal $\theta_j$ as the {\em best}, and the one with maximal $X_j$ as the {\em winner}. Following the convention in the ranking and selection literature, we assume that if there are multiple largest $\theta_j$, then one is arbitrarily marked as the best. Note that in cases where it is more interesting to ask which is the smallest population --- for example, if $X_j$ is the number of patients on treatment $j$ who suffer a heart attack during a trial --- we can change variables to $-X$.

Write the order statistics of $X$ as
\[
X_{[1]} \geq X_{[2]} \geq \cdots \geq X_{[n]},
\]
and define $\theta_{[j]}$ as the entry of $\theta$ corresponding to the $j$th order statistic of $X$ (so $\theta_{[1]}$ might {\em not} equal $\max_j \theta_j$, for example). We assume any ties between entries of $X$ are broken randomly.

@@

Generalizing the procedures described in Section~\ref{sec:iowa} we obtain the following:
\begin{enumerate}
\item {\em Is the winner really the best?} Carry out the unadjusted pairwise test comparing the winner to the runner-up. If the test rejects at level $\alpha$, conclude that the winner is really the best.
\item {\em By how much?} Construct the unadjusted pairwise confidence interval comparing the winner to the runner-up, and report the lower confidence bound obtained for $\theta_{[1]}-\theta_{[2]}$.
\item {\em Is the runner-up really the second best, etc.?} Continue by comparing the runner-up to the second-runner-up, again using the unadjusted pairwise test, and so on down the list comparing adjacent values. Stop the first time the test accepts.
\end{enumerate}

We now state our main theorem: under a mild technical assumption, Procedures 1--3 described above are statistically valid, even accounting for the selection.
\begin{theorem}\label{thm:main}
Assume the model~\eqref{eq:expfam} holds and $g(x)$ is a Schur-concave function. Then:
\begin{enumerate}
\item Procedure 1 has level $\alpha$ conditional on the best population not winning (otherwise an error is impossible).
\item Procedure 2 gives an exact $1-\alpha$ lower confidence bound for $\theta_{[1]} - \max_{j \neq [1]} \theta_{j}$.
\item Procedure 3 is a conservative stepdown procedure with familywise error rate (FWER) no larger than $\alpha$.
\end{enumerate}
\end{theorem}

We define Schur-concave functions and discuss their properties in Section~\ref{sec:maj}. Because any log-concave and symmetric function is Schur-concave, Theorem~\ref{thm:main} applies to all of the cases discussed above. The proof relies on the conditional selective-inference framework of~\citet{Fithian:2014ws} combined with classical multiple-testing methods, as well as ideas involving majorization and Schur-concavity.

Note that these procedures make an implicit adjustment for selection because they use two-tailed unadjusted tests. If we instead based our tests on an independent realization $X^*$, then (for example) Procedure 1 could use a right-tailed version of the unadjusted pairwise test. In the case $n=2$ Procedure 1 amounts to a simple two-tailed test of the null hypothesis $\theta_1=\theta_2$, and it is intuitively clear that a one-tailed test would be anti-conservative. More surprising is that, no matter how large $n$ is, Procedures 1--3 require no further adjustment beyond what is required when $n=2$.

% Informally, this can possibly be explained by noting that for large $n$, $X_{[2]}$ suffers from a positive selection bias --- a ``runner-up's curse,'' so to speak --- which is nearly as severe as the bias of $X_{[1]}$.

% \begin{enumerate}

% \item This has a hidden upward selection bias, or ``winner's curse'', as we are only testing this hypothesis as Trump is leading in the polls. In other words, the hypothesis itself is random. Alternatively, we can note that the marginal type I error rate is not quite straightforward. Specifically, the null hypothesis is ``$H_0 \left(X\right): \text{Leader in polls has most support}$'', which is dependent on random poll data $X$ through $j^*\left(X\right)$, the leader in polls. Then for any distribution $F$ the marginal type I error rate is

% \begin{equation}
% \PP_F\left(\text{reject } H_0\right) = \sum_{\substack{j \text{ is a candidate} \\ j \text{ does not have most support}}} \PP_F\left(\text{reject } H_0 \middle| j \text{ leads in polls}\right) \PP_F \left(j \text{ leads in polls}\right).
% \label{eqn:marginal_error}
% \end{equation}

% \item Comparing Trump to all other candidates involves multiple comparisons. Without considering the structure of this problem, common methods that perform multiple comparisons may largely decrease the test power.
% \end{enumerate}

% We will construct a test as outlined by Fithian, Sun and Taylor in \cite{Fithian:2014ws}. The fundamental idea is to consider the selective type I error rate
% $$\PP_F\left(\text{reject } H_0 \middle| j \text{ leads in polls} \right), ~~~~~~~~ \text{where $j$ is a candidate,}$$
% instead. Note that controlling this selective type I error rate at $\alpha$ provides control of the marginal type I error rate at $\alpha$ too. From \Cref{eqn:marginal_error},
% \begin{align}
% \PP_F\left(\text{reject } H_0\right) & = \sum_{\substack{j \text{ is a candidate} \\ j \text{ does not have most support}}} \PP_F\left(\text{reject } H_0 \middle| j \text{ leads in polls}\right) \PP_F \left(j \text{ leads in polls}\right). \nonumber \\
% & \le \sum_{\substack{j \text{ is a candidate} \\ j \text{ does not have most support}}} \alpha \PP_F \left(j \text{ leads in polls}\right). \nonumber \\
% & = \alpha \PP\left(\text{Leader in polls does not have most support}\right) \nonumber \\
% & \le \frac{n-1}{n} \alpha. \label{eqn:selective_vs_marginal}
% \end{align}

% Surprisingly, this also tackles the second issue above. The test reduces the multiple comparison problem into one comparison between Trump and Cruz. In fact, if 100 John Does entered the race with practically no support from the Iowa Republicans, the change in number of candidates will not change the test.

% More generally, when we observe a random vector $(X_1, \ldots, X_n)$ from an exponential family, and whichever random variable is largest, we want to test whether it is in fact stochastically larger than the other random variables.

@@

Gutmann and Maymin \cite{Gutmann:1987fk} have worked on a similar question where the observations are from a location family, and the location parameters are compared. In contrast, a related paper by Karnnan and Panchapakesan \cite{Karnnan:2009iv} discussed a scale family instead, comparing the scale parameters. Bofinger \cite{Bofinger:1991hv} gave a location-scale family generated by the exponential distribution as well. However, as location-scale families have limited overlap with exponential families, the counterpart question for exponential families have remained unaddressed. Location-scale families assumed independence between observations, while this is not required in exponential families. A follow-up paper by Maymin and Gutmann \cite{Maymin:1992fz} gave a slightly more general result, applicable to more than location-scale families, yet still under the constraints of independence.

In the vicinity of the polling problem above, Nettleton \cite{Nettleton:2009ht} provided an asymptotic test for multinomial distribution, imposing the selection by considering intersection-union tests. Ng and Panchapakesan \cite{Ng:2007cn} also analyzed the multinomial case for an exact test, but with an additional restriction that maximum observation is a known constant.

The subset selection problem solved in \cite{Gupta:1967wg} by Gupta and Nagel can indeed be used to solve the multinomial case. They selected a subset that contains the largest, or ``tagged'', parameter with probability $1-\alpha$. Confirming the largest parameter is thus equivalent to asserting the selected subset has size $1$. \Cref{sec:multinomial_example} includes a comparison of our proposed test and their test.

% \subsection{Main result}

% In this paper, we focus on Schur-concave exponential family, a broad class of distribution that is defined in \Cref{sec:exponential_families} and includes various examples listed in \Cref{sec:applications}. We constructed a selective level-$\alpha$ test for verifying that the largest observation is stochastically largest as well. The test provides selective type I error control, as well as enables us to construct confidence bound on the differences between the natural parameters and verify the order of the rest of the natural parameters. \WFcomment{Still feels a bit abstract when you discuss these goals here.} \KHcomment{What if we don't mention the side results at all?}

\subsection{Outline}

\Cref{sec:multinomial_example} gives an exploratory example based on the multinomial distribution, along with a comparison to an earlier method by Gupta and Nagel. \Cref{sec:exponential_families} gives an in-depth discussion and proof of the main result. \Cref{sec:selective_confidence_bound,sec:step_down_procedure} extend the main result to construct confidence bounds for the gap between natural parameters and a step-down procedure that verifies the order of the rest of the natural parameters with FWER and FDR control, respectively. \Cref{sec:applications} provides a few applications, including both distributions with known tests and those without.

\section{Multinomial example}
\label{sec:multinomial_example}

Suppose we observe
$$\left(X_1, \ldots, X_n\right) \sim \text{Multinomial}\left(m, \pi\right)$$
and we want to test
$$H_{0j}: \pi_j \le \max_{k \ne j} \pi_k = \bigcup_{k \ne j} H_{0jk}: \pi_j \le \pi_k$$
when $X_j$ is larger than all other $X_k$. In case of ties, we break them independently and randomly. The distribution, written in an exponential family format, and up to constant factors, is
$$\exp\left(X_1 \log \pi_1 + \cdots + X_n \log \pi_n\right) \frac{1_{X_1 + \cdots + X_n = m}}{X_1! \cdots X_n!}.$$

Without loss of generality, we will assume $X_1$ is considered largest. Following the construction in \cite{Fithian:2014ws}, if we want to construct $p$-values $p_{1j}$ for $H_{01j}$, we need to first re-parametrize such that the selection event $A_1$ can be expressed in terms of the sufficient statistic. The distribution is now
$$\exp\left(\frac{X_1 - X_j}{2} \left(\log \pi_1 - \log \pi_j\right) + \cdots + \frac{X_1 + X_j}{2} \left(\log \pi_1 + \log \pi_j\right) + \cdots + X_n \log \pi_n\right) \frac{1_{X_1 + \cdots + X_n = m}}{X_1! \cdots X_n!}.$$

Next we condition on the sufficient statistics for the nuisance parameter, in other words, we consider the conditional law
$$\mathcal{L}_{\pi_1 = \pi_j}\left(\frac{X_1 - X_j}{2} \middle| X_2, \ldots, \frac{X_1 + X_j}{2}, \ldots, X_n, A_1\right),$$
or equivalently, the conditional law
$$\mathcal{L}_{\pi_1 = \pi_j}\left(X_1 \middle| X_2, \ldots, \frac{X_1 + X_j}{2}, \ldots, X_n, A_1\right),$$
which is the distribution $\text{Binomial}\left(X_1 + X_j, \frac{1}{2}\right)$ truncated by the event $A_1$. The one-sided $p$-value $p_{1j}$ is thus given as
\begin{align*}
p_{ij}\left(x_1, \ldots, x_n\right) & = \PP\left(X_1 \ge x_1 \middle| X_2 = x_2, \ldots, \frac{X_1 + X_j}{2} = \frac{x_1 + x_j}{2}, \ldots, X_n = x_n, A_1\right) \\
& = \frac{\sum_{t = x_1}^{x_1 + x_j} \binom{x_1 + x_j}{t} \left(\frac{1}{2}\right)^{x_1 + x_j}}{\sum_{t = \left(x_1 + x_j\right)/2}^{x_1 + x_j} \binom{x_1 + x_j}{t} \left(\frac{1}{2}\right)^{x_1 + x_j}} \\
& = \frac{\sum_{t = x_1}^{x_1 + x_j} \binom{x_1 + x_j}{t}}{\sum_{t = \left(x_1 + x_j\right)/2}^{x_1 + x_j} \binom{x_1 + x_j}{t}},
\end{align*}
and to account for the tie-breaking we adopted earlier, we
\begin{itemize}
\item multiply the first term in the numerator and the denominator by $1/k$ where $k$ is the number of observations tied for largest; otherwise,
\item multiply the first term in the denominator by $1/2$ if $X_1$ and $X_j$ has the same parity, as the sum starts $X_1 = X_j = \frac{x_1 + x_j}{2}$.
\end{itemize}
Noticing that if $X_1, \ldots, X_j$ all tied the $p$-value is $1$, and the denominator is really $1/2$ otherwise \WFcomment{I don't think that's true for $j \neq 2$: say $X=(100, 99, 10)$. Then the denominator is }, we can write
$$p_{1j} = 1 \wedge 2 \sum_{t=X_1}^{X_1 + X_j} \binom{X_i + X_j}{t}.$$

It is possible to check that $p_{12} \ge p_{1j}$ for any $j > 2$. The type I error rate is
$$\PP_{H_{01}}\left(p_{12} < \alpha \middle| A_1\right) \stackrel{\text{(i)}}{=} \max_{j>1} \PP_{H_{01j}}\left(p_{12} < \alpha \middle| A_1\right) \stackrel{\text{(ii)}}{\le} \max_{j>1} \PP_{H_{01j}}\left(p_{1j} < \alpha \middle| A_1\right) \stackrel{\text{(iii)}}{\le} \alpha,$$
where (i) follows from $H_{01} = \bigcup_{j>1} H_{01j}$; (ii) is true as $p_{12} \ge p_{1j}$ as long as $X_1 \ge X_2 \ge X_j$; (iii) follows from the construction of $p_{1j}$. \KHcomment{I think the probability in the middle (between (i) and (ii)) is well-defined, as $p_{12}$ is just a random variable dependent on $X$.} Thus it is sufficient to construct a one-sided test between $X_1$ and $X_2$ and consider only $p_{12}$.

It may appear quite surprising that ``verifying a winner'' as we propose in this section requires no adjustment for multiplicity, in the sense that we only need to compare the winner to the runner-up, just as we would have done if $n = 2$. When $n=2$, the observation $X_1$ follows Binomial$\left(X_1 + X_2, \frac{1}{2}\right)$, coinciding with the test we proposed. After all, we know that the winner suffer’s from a winner's curse: given $X_5$ is the winner, $X_5 / m$ would overestimate $\pi_5$, and the larger $n$ is, the more severe this bias is likely to be. However, if $X_9$ is the runner-up, then $X_9 / m$ is likely biased upward nearly as much as $X_5$. Informally, we might say that $X_9$ suffer’s from a ``runner-up's curse'' that is nearly as bad as the winner's curse; as a result, $X_5 - X_9$ is actually not all that biased upward. By multiplying the $p$-value by $2$, or equivalently performing a two-tailed test, we take care of what little bias there is. Indeed, performing a two-tailed test can be viewed as a sort of ``correction for multiplicity'', albeit one that we would have to make even if $n = 2$.

\subsection{Connection with the subset selection problem}
\label{sec:compare_subset}

\Cref{fig:power} gives the power curves for $\text{Multinomial}\left(m, \pi\right)$ and
$$\pi \propto \left(e^\delta, 1, \ldots, 1\right),$$
for various combinations of $m$ and $n$. For Gupta and Nagel, we use $\alpha = 0.05$; but in light of the extra factor of $\frac{n-1}{n}$ in \Cref{eqn:selective_vs_marginal}, we will apply the selective procedure with $\frac{n}{n-1} \alpha$ such that the marginal type I error rate of both procedures are controlled at $\alpha$. \WFcomment{Can we justify this? Our procedure has error $\alpha$ in the case where all $\pi_j$ are equal.} \KHcomment{The selective error rate is $\alpha$, but Gupta and Nagel really only had marginal error rate. If all $\pi_j$ are equal, the marginal error rate is $\frac{n-1}{n} \alpha$.}  \WFcomment{Yes OK I think you're right. Because for G \& N the ``marginal error rate'' is defined by arbitrarily marking one of the tied populations as the best and then asking whether that marked one gets into the confidence set.} Gupta and Nagel coincides with our test at $n = 2$; however as $n$ grows, the selective test shoes significantly more power than Gupta and Nagel's test.

\section{Exponential families}
\label{sec:exponential_families}

We would like to extend the result from multinomials to exponential families. We henceforth focus on exponential distribution taking the form
\begin{equation}
X \sim \exp\left(\theta_1 X_1 + \cdots + \theta_n X_n\right) g\left(X\right) h\left(\theta\right),
\label{eqn:exp_family}
\end{equation}
defined on the Lebesgue or counting measure. It is reasonable to assume that the family is ``symmetric'', i.e.\ if $\theta_j = \theta_k$, then the probability distribution remains the same when $X_j$ and $X_k$ are swapped. This corresponds to requiring $g\left(X\right)$ to be symmetric. We would also like to introduce a fairly mild additional condition that $g$ is {\em Schur-concave}. But first we recall the notion of {\em majorization}, defined on both $\mathbb{R}^n$ and $\mathbb{Z}^n$. \KHcomment{We are making those assumptions. For Lebesgue, there should be no atoms. For counting measure, they should all line up perfectly. I don't think this result generalizes very well for other measures.}

\begin{definition}
For two vectors $a$ and $b$ in $\RR^n$, suppose permuting the two vectors in descending order gives
$a_{\left(1\right)} \ge \cdots \ge a_{\left(n\right)}$ and $b_{\left(1\right)} \ge \cdots \ge b_{\left(n\right)}$. We say that $a \succeq b$ ($a$ majorizes $b$) if for $1 \le i < n$,
\begin{align*}
a_{\left(1\right)} + \cdots + a_{\left(i\right)} & \ge b_{\left(1\right)} + \cdots + b_{\left(i\right)} \\
a_{\left(1\right)} + \cdots + a_{\left(n\right)} & = b_{\left(1\right)} + \cdots + b_{\left(n\right)}
\end{align*}
This forms a partial order in $\RR^n$.
\end{definition}

Intuitively, majorization is a partial order that monitors the evenness of a vector: the more even a vector is, the ``smaller'' it is. There are two properties of majorization that we will use in this section.

\begin{lemma}
\label{lma:two_properties}
\begin{itemize}
\item Suppose $\left(x_1, x_2, x_3, \ldots\right)$ and $\left(x_1, x_2', x_3' \ldots\right)$ are two vectors in $\RR^n$. Then
$$\left(x_1, x_2, x_3, \ldots\right) \succeq \left(x_1, x_2', x_3', \ldots\right) \text{ if and only if } \left(x_2, x_3, \ldots\right) \succeq \left(x_2', x_3', \ldots\right).$$
\item (Principle of transfer) If $x_1 > x_2$ and $t \ge 0$, then
$$\left(x_1 + t, x_2, x_3, \ldots\right) \succeq \left(x_1, x_2 + t, x_3\right).$$
If the $t \le 0$, the majorization is reversed.
\end{itemize}
\end{lemma}

\begin{proof}
\begin{itemize}

\item The property follows from an equivalent formulation of majorization listed in Marshall, Olkin and Arnold \cite{Marshall:2010hb}, where $x \succeq y$ if and only if
$$\sum_{j=1}^n x_n = \sum_{j=1}^n y_n ~~~~ \text{ and } ~~~~ \sum_{j=1}^n \left(x_j - a\right)_+ \ge \sum_{j=1}^n \left(y_j - a\right)_+ \text{ for all } a \in \RR.$$

\item Proved in Marshall, Olkin and Arnold.

\end{itemize}
\end{proof}

@@

A Schur-concave function is symmetric by default since $a \succeq b$ and $b \succeq a$ if and only if $b$ is a permutation of the coordinates of $a$. Conversely a symmetric and log-concave function is Schur-concave \cite{Marshall:2010hb}, thus many common exponential family distributions lie in this class. Schur-concave function indeed appears a lot, \Cref{sec:applications} includes both continuous and discrete examples.

Now suppose the carrier distribution $g$ in \Cref{eqn:exp_family} is indeed Schur-concave. We hereon call the class of such distribution satisfying this {\em Schur-concave exponential family}.

>>>>

\captionsetup{width=0.8\textwidth}
\DeclareMathOperator*{\argmax}{arg\,max}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{lemma}[corollary]{Lemma}
\newtheorem{theorem}[corollary]{Theorem}
\setlength{\tabcolsep}{1em}
\setlist{listparindent=\parindent, parsep=0pt}

\newcommand{\EE}{\mathbb{E}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}

@@

\title{Rank Verification for Exponential Families}
\author{Kenneth Hung \and William Fithian}
\date{\today}
\maketitle

@@

\section{Introduction}
\label{sec:intro}

\subsection{Motivating Example: Iowa Republican Caucus Poll}
\label{sec:iowa}

\Cref{tbl:poll} shows the result of a Quinnipiac University poll asking 890 Iowa Republicans their preferred candidate for the Republican presidential nomination \citep{quinnipiac}. Donald Trump led with $31\%$ of the vote, Ted Cruz came second with $24\%$, Marco Rubio third with $17\%$, and ten other candidates including ``Don't know'' trailed behind.

@@

Seeing that Trump leads this poll, several salient questions may occur to us: Is Trump really winning, and if so by how much? And analogously, is Cruz really in second, is Rubio really in third, and so on? Note that there is implicitly a problem of multiple comparisons here, because if Cruz had led the poll instead, we would be asking a different set of questions. Indeed, the selection issue appears especially pernicious due to the so-called ``winner's curse'': given that Trump leads the poll, it more likely than not overestimates his support.

Nevertheless, if we blithely ignore the selection issue, we might carry out the following analyses to answer the questions we posed before at significance level $\alpha = 0.05$. We assume for simplicity that the poll represents a simple random sample of Iowa Republicans, i.e., that the data are a multinomial sample of size $890$ and underlying probabilities $\left(\pi_{\text{Trump}}, \pi_{\text{Cruz}}, \ldots\right)$.\footnote{The reality is a bit more complicated: before releasing the data, Quinnipiac has processed it to make the reported result more representative of likely caucus-goers.}

\begin{enumerate}
\item {\em Is Trump really winning?} If Trump and Cruz were in fact tied, then Trump's share of their combined 490 votes would be distributed as $\text{Binom}\left(490, 0.5\right)$. Because the (two-tailed) $p$-value for this pairwise test is $p = 0.006$, we reject the null and conclude that Trump is really winning.

\item {\em By how much?} Using an exact $95\%$ interval for the same binomial model, we conclude Trump has at least $7.5\%$ more support than Cruz (i.e., $\pi_{\text{Trump}} \ge 1.075 \,\pi_{\text{Cruz}}$) and also leads the other candidates by at least as much.

\item {\em Is Cruz in second, Rubio in third, etc.?} We can next compare Cruz to Rubio just as we compared Trump to Cruz (again rejecting because $214$ is significantly more than half of 365), then Rubio to Carson, and so on, continuing until we fail to reject. The first four comparisons are all significant at level $0.05$, but Paul and Bush are tied so we stop.
\end{enumerate}

@@

\subsection{Generic Problem Setting and Main Result}
\label{sec:probset}

Generically, we will consider data drawn from an exponential family model with density
\begin{equation}
X \sim \exp\left(\theta'x - \psi\left(\theta\right)\right) g\left(x\right),
\tag{*}
\label{eqn:expfam}
\end{equation}
with respect to either the Lebesgue on $\RR^n$ or counting measure on $\ZZ^n$. We assume further that $g\left(x\right)$ is symmetric. In addition to the multinomial family, model~\eqref{eqn:expfam} also encompasses settings such as comparing independent binomial treatment outcomes in a clinical trial, competing sports teams under a Bradley--Terry model, entries of a Dirichlet distribution, and many more; see \Cref{sec:examples} for these and other examples.

We will generically use the term {\em population} to refer to the treatment group, sports team, political candidate, etc. represented by a given random variable $X_j$. As we will see, $\theta_j \ge \theta_k$ if and only if $X_j$ is stochastically larger than $X_k$; thus, there is a well-defined stochastic ordering of the populations that matches the ordering of the entries of $\theta$. We will refer to the population with maximal $\theta_j$ as the {\em best}, the one with maximal $X_j$ as the {\em winner}, and the one with maximal $X_j$ other than the winner as the {\em runner-up}. Following the convention in the ranking and selection literature, we assume that if there are multiple largest $\theta_j$, then one is arbitrarily marked as the best. Note that in cases where it is more interesting to ask which is the smallest population --- for example, if $X_j$ is the number of patients on treatment $j$ who suffer a heart attack during a trial --- we can change variables to $-X$.

Write the order statistics of $X$ as
$$X_{[1]} \ge X_{[2]} \ge \cdots \ge X_{[n]},$$
where $[j]$ will denote the random index for the $j$-th order statistic. Define $\theta_{[j]}$ as the entry of $\theta$ corresponding to the $j$-th order statistic of $X$ (so $\theta_{[1]}$ might {\em not} equal $\max_j \theta_j$, for example). We assume any ties between entries of $X$ are broken randomly.

@@

Generalizing the procedures described in \Cref{sec:iowa} we obtain the following:
\begin{enumerate}
\item {\em Is the winner really the best?} Carry out the unadjusted pairwise test comparing the winner to the runner-up. If the test rejects at level $\alpha$, conclude that the winner is really the best.
\item {\em By how much?} Construct the unadjusted pairwise confidence interval comparing the winner to the runner-up, and report the lower confidence bound obtained for $\theta_{[1]}-\theta_{[2]}$ if it were nonnegative, report $-\infty$ otherwise.
\item {\em Is the runner-up really the second best, etc.?} Continue by comparing the runner-up to the second-runner-up, again using the unadjusted pairwise test, and so on down the list comparing adjacent values. Stop the first time the test accepts.
\end{enumerate}
Here Procedure $2$ is modified due to technicalities. A less extreme modification is provided in this paper later.

We now state our main theorem: under a mild technical assumption, Procedures 1--3 described above are statistically valid, even accounting for the selection.
\begin{theorem}
Assume the model~\eqref{eqn:expfam} holds and $g\left(x\right)$ is a Schur-concave function. Then:
\begin{enumerate}
\item Procedure 1 has level $\alpha$ conditional on the best population not winning (otherwise an error is impossible).
\item Procedure 2 gives a conservative $1-\alpha$ lower confidence bound for $\theta_{[1]} - \max_{j \ne [1]} \theta_{j}$.
\item Procedure 3 is a conservative step-down procedure with familywise error rate (FWER) no larger than $\alpha$.
\end{enumerate}
\label{thm:main}
\end{theorem}

We define Schur-concave functions and discuss their properties in \Cref{sec:maj}. Because any log-concave and symmetric function is Schur-concave, \Cref{thm:main} applies to all of the cases discussed above. The proof relies on the conditional selective-inference framework of \citet{Fithian:2014ws} combined with classical multiple-testing methods, as well as ideas involving majorization and Schur-concavity.

Note that these procedures make an implicit adjustment for selection because they use two-tailed unadjusted tests. If we instead based our tests on an independent realization $X^*$, then, for example, Procedure 1 could use a right-tailed version of the unadjusted pairwise test. In the case $n = 2$ Procedure 1 amounts to a simple two-tailed test of the null hypothesis $\theta_1 = \theta_2$, and it is intuitively clear that a one-tailed test would be anti-conservative. More surprising is that, no matter how large $n$ is, Procedures 1--3 require no further adjustment beyond what is required when $n = 2$.

@@

\citet{Gutmann:1987fk} have worked on a similar question where the observations are from a location family, and the location parameters are compared. In contrast, a related paper by \citet{Karnnan:2009iv} discussed a scale family instead, comparing the scale parameters. \citet{Bofinger:1991hv} gave a location-scale family generated by the exponential distribution as well. However, as location-scale families have limited overlap with exponential families, the counterpart question for exponential families have remained unaddressed. Location-scale families assumed independence between observations, while this is not required in exponential families. A follow-up paper by \citet{Maymin:1992fz} gave a slightly more general result, applicable to more than location-scale families, yet still under the constraints of independence.

In the vicinity of the polling problem above, \citet{Nettleton:2009ht} provided an asymptotic test for multinomial distribution, imposing the selection by considering intersection-union tests. \citet{Ng:2007cn} also analyzed the multinomial case for an exact test, but with an additional restriction that maximum observation is a known constant.

The subset selection problem solved in \citet{Gupta:1967wg} can indeed be used to solve the multinomial case. They selected a subset that contains the largest, or ``tagged'', parameter with probability $1-\alpha$. Confirming the largest parameter is thus equivalent to asserting the selected subset has size $1$. \Cref{sec:subsetsel} includes a comparison of our proposed test and their test.

\subsection{Outline}
\Cref{sec:maj} gives a formal definition of the condition we are imposing, as well as common examples satisfying the condition. \Cref{sec:winnerbest} proves that an unadjusted pairwise test comparing the winner to the runner-up is valid, as well as a demonstration of the power of this test. This section also explains the validity of level $\alpha$ conditional on the best population not winning. \Cref{sec:confbound,sec:stepdown} prove Procedure 2 and 3 are valid. We will finally examine a published study with our methods in \Cref{sec:conclusion}.

\section{Majorization and Schur-concavity}
\label{sec:maj}

We start by reviewing the notion of {\em majorization}, defined on both $\RR^n$ and $\ZZ^n$.

\begin{definition}
For two vectors $a$ and $b$ in $\RR^n$ (or $\ZZ^n$), suppose permuting the two vectors in descending order gives
$a_{\left(1\right)} \ge \cdots \ge a_{\left(n\right)}$ and $b_{\left(1\right)} \ge \cdots \ge b_{\left(n\right)}$. We say that $a \succeq b$ ($a$ majorizes $b$) if for $1 \le i < n$,
\begin{align*}
a_{\left(1\right)} + \cdots + a_{\left(i\right)} & \ge b_{\left(1\right)} + \cdots + b_{\left(i\right)} \\
a_{\left(1\right)} + \cdots + a_{\left(n\right)} & = b_{\left(1\right)} + \cdots + b_{\left(n\right)}
\end{align*}
This forms a partial order in $\RR^n$ (or $\ZZ^n$).
\end{definition}

Intuitively, majorization is a partial order that monitors the evenness of a vector: the more even a vector is, the ``smaller'' it is. There are two properties of majorization that we will use in the proofs.

\begin{lemma}
\begin{enumerate}
\item Suppose $\left(x_1, x_2, x_3, \ldots\right)$ and $\left(x_1, x_2', x_3' \ldots\right)$ are two vectors in $\RR^n$. Then
$$\left(x_1, x_2, x_3, \ldots\right) \succeq \left(x_1, x_2', x_3', \ldots\right) \text{ if and only if } \left(x_2, x_3, \ldots\right) \succeq \left(x_2', x_3', \ldots\right).$$
\item (Principle of transfer) If $x_1 > x_2$ and $t \ge 0$, then
$$\left(x_1 + t, x_2, x_3, \ldots\right) \succeq \left(x_1, x_2 + t, x_3\right).$$
If the $t \le 0$, the majorization is reversed.
\end{enumerate}
\label{lma:twoprop}
\end{lemma}

\begin{proof}
\begin{enumerate}
\item The property follows from an equivalent formulation of majorization listed in \citet{Marshall:2010hb}, where $x \succeq y$ if and only if
$$\sum_{j=1}^n x_n = \sum_{j=1}^n y_n ~~~~ \text{ and } ~~~~ \sum_{j=1}^n \left(x_j - a\right)_+ \ge \sum_{j=1}^n \left(y_j - a\right)_+ \text{ for all } a \in \RR.$$

\item Proved in \citet{Marshall:2010hb}. \qedhere
\end{enumerate}
\end{proof}

@@

A Schur-concave function is symmetric by default since $a \succeq b$ and $b \succeq a$ if and only if $b$ is a permutation of the coordinates of $a$. Conversely a symmetric and log-concave function is Schur-concave \cite{Marshall:2010hb}, thus many common exponential family distributions lie in this class. Schur-concave functions indeed appear a lot, \Cref{sec:examples} includes both continuous and discrete examples.

\Cref{thm:main} requires that $g\left(x\right)$ in model~\eqref{eqn:expfam} to be Schur-concave. It was stated in the ensuing discussion that this condition is mild, and that the examples are abundant. To name a few:

\subsection{Examples}
\label{sec:examples}

\begin{enumerate}
\item Independent binomial treatment outcome in a clinical trial. If each of the $n$ different treatments are applied $m$ patients independently, the number of positive outcomes $X_j$ for treatment $j$ can be modeled by $\text{Binomial}\left(m, p_j\right)$. The best treatment would be the treatment with the highest success probability $p_j$. The joint distribution of $X$ is given by
$$p\left(x\right) \propto \exp\left(x' \left(\log p\right)\right) \frac{1}{x_1! \left(m-x_1\right)! \cdots x_n! \left(m-p_x\right)!}$$
The carrier measure above is Schur-concave.
